---
title: "DE_analysis"
author: "Milena Trabert"
date: "2026-03-27"
output: html_document
---

## load libraries
```{r}
# install all dependencies
if (FALSE){
  if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
  BiocManager::install("DESeq2") # compiled from source 
  BiocManager::install("edgeR")
  # install tidyverse (ggplot2 is in tidyverse)
  install.packages("tidyverse") 
  install.packages("gtable") 
  install.packages("VennDiagram") 
}
library(tidyverse)
library(ggplot2)
library(edgeR)
library(limma)
library(VennDiagram)
library(dplyr)
```

## prepare data with all samples
```{r}
in_filename <- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/gene_counts_standard.txt"
out_norm_counts <- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/gene_counts_normalized.tsv"
metadata <- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/larvae_metadata.csv"

## load data
  count_data <- read.csv(in_filename, header = TRUE, sep = "\t",comment.char = "#", check.names = FALSE)
  count_data <- count_data %>% select(-one_of('Chr','Start','End','Strand','Length')) 
  # make the Geneid the row id 
  count_data <- count_data %>% remove_rownames %>% column_to_rownames(var="Geneid")
## meta data
  DE_metadata_all <- read.csv(metadata, header = TRUE, sep = ",",comment.char = "#")
  # only include metadata when the samples are in the counts
  meta_data <- DE_metadata_all[which((DE_metadata_all$ID %in% names(count_data))==TRUE), ]
  meta_data <- meta_data %>% remove_rownames %>% column_to_rownames(var="ID")
  # reorder so it's in the same order as the count data columns
  meta_data <- meta_data[names(count_data),]
  # check assumptions, should be TRUE
  all(colnames(count_data) %in% rownames(meta_data))
  
## normalize library sizes
  # make sample groups that reflect line and day
  groups <- factor(paste(meta_data$line, meta_data$sex, meta_data$day, sep="_"))
  # read data into data structure
  y <- DGEList(counts=count_data, group=groups)
  #dim(y)
  #[1] 15702    82
  
## filter out low count genes
  # minimum count of 5 in at least some samples, and a minimum total count of 15 across all samples 
  keep <- filterByExpr(y, group=meta_data$line, min.count=5, min.total.count=15)
  y <- y[keep, , keep.lib.sizes=FALSE]
  y$samples$lib.size <- colSums(y$counts) # recompute library sizes
  # dim(y)
  # [1] 11641    82
  
## normalize by library sizes
  y <- calcNormFactors(y)

short_rownames <- c("1-14-M","1-14-F","1-14-M","1-14-M","1-14-F","1-14-M","1-14-M","1-14-M","1-14-M","1-14-F","1-14-M","1-14-M","1-14-M","1-16-M","1-16-M","1-16-M","1-16-F","1-16-M","1-16-F","1-16-M","1-16-F","1-16-M","1-16-M","1-16-M","1-16-M","1-16-M","1-16-M","1-18-M","1-18-M","1-18-F","1-18-F","1-18-M","1-18-M","1-18-M","1-18-M","1-18-M","1-18-F","1-18-F","1-18-F","3-14-M","3-14-M","3-14-F","3-14-M","3-14-M","3-14-F","3-14-M","3-14-F","3-14-M","3-14-M","3-14-F","3-14-M","3-14-F","3-14-F","3-14-M","3-16-F","3-16-M","3-16-F","3-16-M","3-16-M","3-16-F","3-16-M","3-16-F","3-16-F","3-16-M","3-16-M","3-16-M","3-16-F","3-18-F","3-18-M","3-18-M","3-18-M","3-18-M","3-18-M","3-18-M","3-18-M","3-18-F","3-18-M","3-18-F","3-18-M","3-18-M","3-18-M","3-18-M")
png("/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_figures/MDS_males_and_females.png", width = 7, height = 6, units = "in", res = 300)
plotMDS(y, labels = short_rownames, col=as.numeric(y$samples$group))
title("MDS clustering of all samples")
dev.off()
## export normalized counts to file
  log_norm_counts <- cpm(y,normalized.lib.sizes = TRUE,log = TRUE,prior.count = 1)
  write.table(log_norm_counts,file = out_norm_counts, sep = "\t",quote = FALSE)# ,col.names = NA)
  #### !!! manually add first header "geneID" to the file !!! ###
  
```
See Python scripting for the PCA plots


# basic DE with only the males

I will start with DE analysis that involves only the males to get an idea of what is going on. 

## partition and prepare data

```{r} 
## filter count data to only include males 
counts_M <- count_data[,grepl("-M_", colnames(count_data))]
## adjust metadata
meta_data <- DE_metadata_all[which((DE_metadata_all$ID %in% names(counts_M))==TRUE), ]
meta_data <- meta_data %>% remove_rownames %>% column_to_rownames(var="ID")
# reorder so it's in the same order as the count data columns
meta_data <- meta_data[names(counts_M),]
# check assumptions, should be TRUE
all(colnames(counts_M) %in% rownames(meta_data))

# make sample groups that reflect line and day
groups <- factor(paste(meta_data$line, meta_data$day, sep="_"))
# read data into data structure
y <- DGEList(counts=counts_M, group=groups)

## filter out low count genes
# filter by only line group and not day, since we aren't sure how the expression changes between days and we don't want to remove genes that are only rarely expressed
keep <- filterByExpr(y, group=meta_data$line, min.count=5, min.total.count=15)
  y <- y[keep, , keep.lib.sizes=FALSE]
  y$samples$lib.size <- colSums(y$counts) # recompute library sizes
  
# normalize by library sizes
y <- calcNormFactors(y)

# estimate dispersion
MM <- model.matrix(~0 + groups, data = y$samples)
colnames(MM) <- c("SL1_14", "SL1_16", "SL1_18", "SL3_14",  "SL3_16", "SL3_18")
rownames(MM) <- colnames(y)

y <- estimateDisp(y, MM)
plotBCV(y)
short_rownames <- c("1-14","1-14","1-14","1-14","1-14","1-14","1-14","1-14","1-14","1-14","1-16","1-16","1-16","1-16","1-16","1-16","1-16","1-16","1-16","1-16","1-16","1-18","1-18","1-18","1-18","1-18","1-18","1-18","3-14","3-14","3-14","3-14","3-14","3-14","3-14","3-14","3-14","3-16","3-16","3-16","3-16","3-16","3-16","3-16","3-18","3-18","3-18","3-18","3-18","3-18","3-18","3-18","3-18","3-18","3-18","3-18")
png("/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_figures/MDS_males_only.png", width = 7, height = 6, units = "in", res = 300)
plotMDS(y,  labels = short_rownames, col=as.numeric(y$samples$group))
title("MDS clustering of male samples")
dev.off()
```

## Differential expression just males

This is based on Elina's code and the edgeR manual. I will test several contrasts, where each contrast is SL1-SL3 within the same day. Additionally, since we know that day 14 and 16 are similar and 18 is different from both of them, I will specify the within-line contrasts to 18-mean(14,16).
```{r}
fit <- glmFit(y, MM)

my.contrasts <- makeContrasts(
day14 =	SL1_14 - SL3_14, 
day16 =	SL1_16 - SL3_16, 
day18 = SL1_18 - SL3_18,
# compare 18 to the average of 14 and 16 in each line, since 18 is the most different
line_1 = SL1_18 - (SL1_14+SL1_16)/2,
line_3 = SL3_18 - (SL3_14+SL3_16)/2,
levels=MM)

contrast_analysis <- function(y,MM,contrast, table_name, smear_name, smear_ylab, smear_title){
  
  fit <- glmFit(y, MM)
  lrt <- glmLRT(fit, contrast=contrast)
  de <- decideTestsDGE(lrt, p.value=0.05, lfc=1)
  print("glmLRT")
  print(smear_title)
  print(summary(de))
  
  fit <- glmQLFit(y, MM)
  qlf <- glmQLFTest(fit, contrast=contrast)
  de_qlf <- decideTestsDGE(qlf, p.value=0.05, lfc=1)
  print("glmQLFTest")
  print(summary(de_qlf))
  
  print("--------------------------------------------")
  
  ## make output table
  Tags <- topTags(qlf, n=Inf, p.value=0.05)						
  Tags <- Tags$table
  Tags$Gene <- rownames(Tags)
  write.table(Tags, file=table_name, sep="\t")	
  
  ## make smear plot
  de<-decideTestsDGE(qlf, p.value=0.05, lfc=1) 
  detags <- rownames(y)[as.logical(de)]
  png(smear_name, width = 6, height = 4, units = "in", res = 300)
  plotSmear(qlf, de.tags=detags, smearWidth=0.1, smooth.scatter=F, xlim=c(), ylab=smear_ylab, main = smear_title)
  abline(h=c(-1, 1), col="blue")
  dev.off()
  
  output <- list(qlf=qlf,de=de,detags=detags, Tags=Tags)
  return(output)
}

d14_table = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_genes_M_1-3_day14.txt"
d14_smear_name <- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_figures/smear_M_d14.png"
d14_smear_title <- "Male samples day 14"
out_14 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"day14"],table_name=d14_table,smear_name=d14_smear_name,smear_ylab="logFC SL1-SL3",smear_title=d14_smear_title)
qlf14 <- out_14["qlf"]
de14 <- out_14["de"]
detags14 <- out_14["detags"]
Tags14 <- out_14["Tags"]

d16_table = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_genes_M_1-3_day16.txt"
d16_smear_name <- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_figures/smear_M_d16.png"
d16_smear_title <- "Male samples day 16"
out_16 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"day16"],table_name=d16_table,smear_name=d16_smear_name,smear_ylab="logFC SL1-SL3",smear_title=d16_smear_title)
qlf16 <- out_16["qlf"]
de16 <- out_16["de"]
detags16 <- out_16["detags"]
Tags16 <- out_16["Tags"]

d18_table = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_genes_M_1-3_day18.txt"
d18_smear_name <- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_figures/smear_M_d18.png"
d18_smear_title <- "Male samples day 18"
out_18 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"day18"],table_name=d18_table,smear_name=d18_smear_name,smear_ylab="logFC SL1-SL3",smear_title=d18_smear_title)
qlf18 <- out_18["qlf"]
de18 <- out_18["de"]
detags18 <- out_18["detags"]
Tags18 <- out_18["Tags"]

line1_table = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_genes_M_SL1.txt"
line1_smear_name <- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_figures/smear_M_SL1.png"
line1_smear_title <- "Male samples line 1"
out_SL1 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"line_1"],table_name=line1_table,smear_name=line1_smear_name,smear_ylab="logFC d18 - (d14+d16)/2",smear_title=line1_smear_title)
qlfSL1 <- out_SL1["qlf"]
deSL1 <- out_SL1["de"]
detagsSL1 <- out_SL1["detags"]
Tags_SL1 <- out_SL1["Tags"]

line3_table = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_genes_M_SL3.txt"
line3_smear_name <- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_figures/smear_M_SL3.png"
line3_smear_title <- "Male samples line 3"
out_SL3 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"line_3"],table_name=line3_table,smear_name=line3_smear_name,smear_ylab="logFC d18 - (d14+d16)/2",smear_title=line3_smear_title)
qlfSL3 <- out_SL3["qlf"]
deSL3 <- out_SL3["de"]
detagsSL3 <- out_SL3["detags"]
Tags_SL3 <- out_SL3["Tags"]
```

### Venn diagram

General venn diagramm for overlap of differentially expressed genes in either direction between the days
```{r}
library(VennDiagram)

venn_name <- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_figures/DE_days_M_venn.png"
DE_genes <- list(d14 = unlist(detags14), d16 = unlist(detags16), d18 = unlist(detags18))
venn_plot <- venn.diagram(DE_genes, 
                          filename = venn_name, 
                          # output
                          imagetype="png",
                          height = 720,
                          width = 720,
                          resolution = 300,
                          compression = "lzw",
                          # circles
                          fill = c("#DE8F6E", "#2D93AD", "#88AB75"),
                          lwd = 2,
                          lty = 'blank',
                          # Numbers
                          cex = .8,
                          main.fontface = "bold",
                          main.fontfamily = "sans",
                          fontface = "bold",
                          fontfamily = "sans",
                          cat.cex = 0.8,
                          cat.fontface = "bold",
                          cat.default.pos = "outer",
                          cat.pos = c(-27, 27, 135),
                          cat.dist = c(0.055, 0.055, 0.085),
                          cat.fontfamily = "sans",
                          rotation = 1,
                          main="DE genes (q=0.05, lfc >1)",
                          disable.logging = TRUE)

venn_name_lines <- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_figures/DE_lines_M_venn.png"
DE_genes_lines <- list(smallY = unlist(detagsSL1), largeY = unlist(detagsSL3))
venn_plot <- venn.diagram(DE_genes_lines, 
                          filename = venn_name_lines, 
                          # output
                          imagetype="png",
                          height = 720,
                          width = 720,
                          resolution = 300,
                          compression = "lzw",
                          # circles
                          fill = c("#DE8F6E", "#2D93AD"),
                          lwd = 2,
                          lty = 'blank',
                          # Numbers
                          cex = .8,
                          main.fontface = "bold",
                          main.fontfamily = "sans",
                          fontface = "bold",
                          fontfamily = "sans",
                          cat.cex = 0.8,
                          cat.fontface = "bold",
                          cat.default.pos = "outer",
                          cat.pos = c(-27, 27),
                          cat.dist = c(0.03, 0.01),
                          cat.fontfamily = "sans",
                          main="DE genes (q=0.05, lfc >1)",
                          disable.logging = TRUE)
```

## plot sex bias overlap

Plot genes showing DE in any combination of two contrasts

```{r}
plot_DE_overlap <- function(Res1,Res2, sig_p,plot_name, label1, label2, xlabel,ylabel,plot_title){
  
  Tags1 <- topTags(Res1, n=Inf)
  Tags2 <- topTags(Res2, n=Inf)
  abdg <- Tags1$table	
  htg <- Tags2$table
  abdg$gene <- row.names(abdg)
  htg$gene <- row.names(htg)
  
  df <- merge(abdg, htg, by="gene", all=T)
  df <- subset(df, FDR.x<sig_p | FDR.y<sig_p )
  dim(df)
  df$category <- ifelse((df$FDR.x <sig_p) & (df$FDR.y > sig_p | is.na(df$FDR.y)), label1, 
		             ifelse((df$FDR.x > sig_p | is.na(df$FDR.x)) & (df$FDR.y < sig_p), label2, 
		             ifelse((df$FDR.x <sig_p) & (df$FDR.y < sig_p),"Both", 0)))
  
  print(plot_title)
  counts <- df |> count(category)
  print(counts)
  print("-----------------------------------")
  
  abdLogFC <- df[,c("logFC.x")] # Take only the logFC
  htLogFC <- df[,c("logFC.y")]
  abdLogFC <- as.numeric(abdLogFC)
  htLogFC <- as.numeric(htLogFC)
  
  category <- df[,c("category")]
  newDF <- as.data.frame(cbind(abdLogFC, htLogFC))	
  newDF$category <- as.factor(category)

  g = ggplot(data= newDF, aes(x= htLogFC, y= abdLogFC)) + 
    geom_point(aes(colour=factor(category)), alpha=0.6, size=1.75) + 
   	scale_color_manual(values = c("#F2933A", "#4d7298", "#b82946"),name ="Differential expression (q<0.05)", 
                                  labels=c(label1, "both", label2)) +
    xlab(xlabel) + ylab(ylabel) + 
   	ggtitle(plot_title) + 
   	geom_hline(aes(yintercept=0), colour="#990000", linetype="dashed") + 
   	geom_vline(aes(xintercept=0), colour="#990000", linetype="dashed") +  
   	geom_hline(aes(yintercept=1), colour="darkgrey", linetype ="dashed") + 
   	geom_hline(aes(yintercept=-1), colour="darkgrey", linetype ="dashed") + 
   	geom_vline(aes(xintercept =1), colour="darkgrey", linetype ="dashed") + 
   	geom_vline(aes(xintercept =-1), colour="darkgrey", linetype="dashed") +
   	theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
    panel.background = element_blank(), axis.line = element_line(colour = "black")) 
  ggsave(filename = plot_name, width = 6, height = 4, units = "in", plot = g)
}
```

DE between SL1 vs. SL3 and DE in both, for day 14 and day 16
```{r}
day14_16_name <- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_figures/DE_line_overlap_day14_16.png"
plot_DE_overlap(Res1 = qlf14[[1]], Res2 = qlf16[[1]], sig_p = 0.05, plot_name=day14_16_name,
                label1 = "day 14", label2 = "day 16", 
                xlabel = "LFC SL1-SL3 day 14",ylabel = "LFC SL1-SL3 day 16" ,
                plot_title = "differential expression between SL1 and SL3")

day14_18_name <- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_figures/DE_line_overlap_day14_18.png"
plot_DE_overlap(Res1 = qlf14[[1]], Res2 = qlf18[[1]], sig_p = 0.05, plot_name=day14_18_name,
                label1 = "day 14", label2 = "day 18", 
                xlabel = "LFC SL1-SL3 day 14",ylabel = "LFC SL1-SL3 day 18" ,
                plot_title = "differential expression between SL1 and SL3")

day16_18_name <- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_figures/DE_line_overlap_day16_18.png"
plot_DE_overlap(Res1 = qlf16[[1]], Res2 = qlf18[[1]], sig_p = 0.05, plot_name=day16_18_name,
                label1 = "day 16", label2 = "day 18", 
                xlabel = "LFC SL1-SL3 day 16",ylabel = "LFC SL1-SL3 day 18" ,
                plot_title = "differential expression between SL1 and SL3")
```

DE between mean(day14,day16) and day 18, for SL1 and SL3
```{r}
lines_name <- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_figures/DE_day_overlap_SL1_SL3.png"
plot_DE_overlap(Res1 = qlfSL1[[1]], Res2 = qlfSL3[[1]], sig_p = 0.05, plot_name=lines_name,
                label1 = "SL1 (small Y)", label2 = "SL3 (large Y)", 
                xlabel = "LFC d18-mean(d14,d16) SL1",ylabel = "LFC d18-mean(d14,d16) SL3" ,
                plot_title = "differential expression between d18 and mean(d14,d16)")
```


```{r}

```


