---
title: "DE_analysis"
author: "Milena Trabert"
date: "2026-03-27"
output: html_document
---

# Make LogFC tables for all contrasts and data partitions

I will separate the data three times: by sex, line and day to avoid having to fit a three-way interaction.

### load libraries
```{r}
# install all dependencies
if (FALSE){
  if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
  BiocManager::install("DESeq2") # compiled from source 
  BiocManager::install("edgeR")
  # install tidyverse (ggplot2 is in tidyverse)
  install.packages("tidyverse") 
  install.packages("gtable") 
  install.packages("VennDiagram") 
}
library(tidyverse)
library(ggplot2)
library(edgeR)
library(limma)
library(VennDiagram)
library(dplyr)
```

## prepare data with all samples

Load either whole gene set, or filtered to only include X or Y genes.

```{r}
# in_filename <- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/gene_counts_standard.txt"
in_filename <- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/gene_counts_standard_merge_yTor.txt"
outdir <- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/"

# in_filename <- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/gene_counts_standard_only_Y.txt"
# outdir <- ("/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_only_Y/"

# in_filename sprintf<- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/gene_counts_standard_only_X.txt"
# outdir <- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_only_X/"
```

Load all data first and compute right library sizes. Only split by sample after this.

```{r}

out_norm_counts <- sprintf("%sgene_counts_normalized_nolog.tsv", outdir)
metadata <- sprintf("%slarvae_metadata.csv", outdir)

## load data
  count_data <- read.csv(in_filename, header = TRUE, sep = "\t",comment.char = "#", check.names = FALSE)
  count_data <- count_data %>% select(-one_of('Chr','Start','End','Strand','Length')) 
  # make the Geneid the row id 
  count_data <- count_data %>% remove_rownames %>% column_to_rownames(var="Geneid")
## meta data
  DE_metadata_all <- read.csv(metadata, header = TRUE, sep = ",",comment.char = "#")
  # only include metadata when the samples are in the counts
  meta_data <- DE_metadata_all[which((DE_metadata_all$ID %in% names(count_data))==TRUE), ]
  meta_data <- meta_data %>% remove_rownames %>% column_to_rownames(var="ID")
  # reorder so it's in the same order as the count data columns
  meta_data <- meta_data[names(count_data),]
  # check assumptions, should be TRUE
  all(colnames(count_data) %in% rownames(meta_data))
  
## normalize library sizes
  # make sample groups that reflect line and day
  groups <- factor(paste(meta_data$line, meta_data$sex, meta_data$day, sep="_"))
  # read data into data structure
  y <- DGEList(counts=count_data, group=groups)
  #dim(y)
  #[1] 15702    82
## filter out low count genes
  # minimum count of 5 in at least some samples, and a minimum total count of 15 across all samples 
  # min.prop=0 so that the yTors stay even though they are only expressed in SL1-M
  #   In general, all Y-exclusive genes should be kept this way since they are absent in all females
  keep <- filterByExpr(y, group=meta_data$line, min.count=5, min.total.count=5 , min.prop=0)
  y <- y[keep, , keep.lib.sizes=FALSE]
  y$samples$lib.size <- colSums(y$counts) # recompute library sizes
  # dim(y)
  # [1] 11641    82
  
## normalize by library sizes
  y <- calcNormFactors(y)

```

### plot MDS

Based on full data, to compare with PCA
```{r}
short_rownames <- c("1-14-M","1-14-F","1-14-M","1-14-M","1-14-F","1-14-M","1-14-M","1-14-M","1-14-M","1-14-F","1-14-M","1-14-M","1-14-M","1-16-M","1-16-M","1-16-M","1-16-F","1-16-M","1-16-F","1-16-M","1-16-F","1-16-M","1-16-M","1-16-M","1-16-M","1-16-M","1-16-M","1-18-M","1-18-M","1-18-F","1-18-F","1-18-M","1-18-M","1-18-M","1-18-M","1-18-M","1-18-F","1-18-F","1-18-F","3-14-M","3-14-M","3-14-F","3-14-M","3-14-M","3-14-F","3-14-M","3-14-F","3-14-M","3-14-M","3-14-F","3-14-M","3-14-F","3-14-F","3-14-M","3-16-F","3-16-M","3-16-F","3-16-M","3-16-M","3-16-F","3-16-M","3-16-F","3-16-F","3-16-M","3-16-M","3-16-M","3-16-F","3-18-F","3-18-M","3-18-M","3-18-M","3-18-M","3-18-M","3-18-M","3-18-M","3-18-F","3-18-M","3-18-F","3-18-M","3-18-M","3-18-M","3-18-M")
png("/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_figures/MDS_males_and_females.png", width = 7, height = 6, units = "in", res = 300)
plotMDS(y, labels = short_rownames, col=as.numeric(y$samples$group))
title("MDS clustering of all samples")
dev.off()
```

Export normalized counts to file. See Python scripting for the PCA plots based on the counts.
```{r}
  log_norm_counts <- cpm(y,normalized.lib.sizes = TRUE,log = FALSE,prior.count = 1)
  write.table(log_norm_counts,file = out_norm_counts, sep = "\t",quote = FALSE)
```


# get expressed genes on sex chromosomes

Only run one at a time! not consecutively

## Y

```{r}
dim(y)
y_genes <- c("gene-402962","gene-402968","gene-402989","gene-424042","gene-424048","gene-424063","gene-424069","gene-424075","gene-424085","gene-424095","gene-371557","gene-371566","gene-371572","gene-371578","gene-371593","gene-371599","gene-371605","gene-371611","gene-371617","gene-371627","gene-371633","gene-371651","gene-371660","gene-371666","gene-371676","gene-371686","gene-371698","gene-371704","gene-371719","gene-371725","gene-371738","gene-371744","gene-371750","gene-371763","gene-371769","gene-371775","gene-371781","gene-371787","gene-371793","gene-371799","gene-371805","yTor-A","yTor-B","gene-371844","yTor-C","gene-371883","gene-371889","gene-371898","gene-371907","gene-371913","gene-371922","gene-371936","gene-371957","gene-371984","gene-372026","gene-372053","gene-372068","gene-372207","gene-372216","gene-372237","gene-372243","gene-372264","gene-372302")
y <- y[rownames(y$counts) %in% y_genes, , keep.lib.sizes=FALSE]
# remove samples that have 0 expression in all
y <- y[, colSums(y$counts) > 0]
y$samples$lib.size <- colSums(y$counts)
dim(y)
print(rownames(y$counts))
```

## X

```{r}
dim(y)
x_genes <- c("gene-306335","gene-306353","gene-306359","gene-306484","gene-306496","gene-306523","gene-306550","gene-306577","gene-306589","gene-306648","gene-306675","gene-306713","gene-306736","gene-306977","gene-306995","gene-307001","gene-307013","gene-307098","gene-307178","gene-307190","gene-307264","gene-307421","gene-307427","gene-307433","gene-307442","gene-307448","gene-307460","gene-307469","gene-307478","gene-307484","gene-307496","gene-307508","gene-307537","gene-307555","gene-307679","gene-307691","gene-307700","gene-307736","gene-307746","gene-308117","gene-308126","gene-308132","gene-308150","gene-308177","gene-308198","gene-308216","gene-308231","gene-308282","gene-308300","gene-308321","gene-308424","gene-308454","gene-308487","gene-308508","gene-308549","gene-308564","gene-308597","gene-308603","gene-308697","gene-308708","gene-308738","gene-308795","gene-308810","gene-308822","gene-308840","gene-308850","gene-308862","gene-308874","gene-308880","gene-308887","gene-308897","gene-308921","gene-308930","gene-308945","gene-308981","gene-308990","gene-309049","gene-309070","gene-309076","gene-309252","gene-309273","gene-309288","gene-309315","gene-309327","gene-309339","gene-309387","gene-309399","gene-309408","gene-309414","gene-309420","gene-309471","gene-309516","gene-309522","gene-309556","gene-309583","gene-309642","gene-309648","gene-309660","gene-309708","gene-309714","gene-309744","gene-309768","gene-309828","gene-309843","gene-309870","gene-309876","gene-309885","gene-309915","gene-309966","gene-309993","gene-309999","gene-310012","gene-310065","gene-310115","gene-310121","gene-310152","gene-310158","gene-310176","gene-310194","gene-310230","gene-310251","gene-310287","gene-310338","gene-310383","gene-310422","gene-310482","gene-310604","gene-310639","gene-310648","gene-310660","gene-310714","gene-310729","gene-310850","gene-310868","gene-310889","gene-310951","gene-310975","gene-311116","gene-311131","gene-311152","gene-311193","gene-311211","gene-311273","gene-311303","gene-311343","gene-311367","gene-311396","gene-311426","gene-311444","gene-311450","gene-311462","gene-311477","gene-311521","gene-311551","gene-311563","gene-311581","gene-311593","gene-311620","gene-311647","gene-311674","gene-311725","gene-311734","gene-311749","gene-311758","gene-311767","gene-311796","gene-311808","gene-311817","gene-311904","gene-311910","gene-311934","gene-311952","gene-311970","gene-312110","gene-312140","gene-312229","gene-312264","gene-312270","gene-312288","gene-312294","gene-312314","gene-312326","gene-312362","gene-312409","gene-312416","gene-312440","gene-312452","gene-312479","gene-312544","gene-312583","gene-312604","gene-312715","gene-312748","gene-312890","gene-312908","gene-312923","gene-312938","gene-312968","gene-312983","gene-313040","gene-313064","gene-313139","gene-313154","gene-313196","gene-313303","gene-313324","gene-313330","gene-313455","gene-313478","gene-313493","gene-313589","gene-313618","gene-314377","gene-314386","gene-314478","gene-314490","gene-314511","gene-314576","gene-314588","gene-314599","gene-314611","gene-314715","gene-314807","gene-314825","gene-314834","gene-314864","gene-314873","gene-314900","gene-314921","gene-314945","gene-314951","gene-314980","gene-315004","gene-315034","gene-315114","gene-315138","gene-315170","gene-315215","gene-315254","gene-315287","gene-315379","gene-315455","gene-315467","gene-315482","gene-315515","gene-315539","gene-315551","gene-315569","gene-315578","gene-315593","gene-315603","gene-315630","gene-315640","gene-315808","gene-315814","gene-315826","gene-315967","gene-315991","gene-316014","gene-316126","gene-316255","gene-316282","gene-316300","gene-316413","gene-316434","gene-316459","gene-316474","gene-316544","gene-316589","gene-316686","gene-316713","gene-316731","gene-316755","gene-316779","gene-316827","gene-316836","gene-316880","gene-316886","gene-316951","gene-317002","gene-317103","gene-317127","gene-317184","gene-317205","gene-317217","gene-317235","gene-317256","gene-317283","gene-317289","gene-317336","gene-317372","gene-317405","gene-317564","gene-317573","gene-317951","gene-317969","gene-318475","gene-320235","gene-320352","gene-320375","gene-320404","gene-320470","gene-320479","gene-320494","gene-320541","gene-320568","gene-320594","gene-320647","gene-320770","gene-320782","gene-320803","gene-320812","gene-320821","gene-320833","gene-320854","gene-320878","gene-320978","gene-320990","gene-321011","gene-321017","gene-321078","gene-321107","gene-321177","gene-321219","gene-321346","gene-321367","gene-321450","gene-321510","gene-321546","gene-321587","gene-321629","gene-321638","gene-321653","gene-321691","gene-321706","gene-321733","gene-321745","gene-321775","gene-321790","gene-321814","gene-321841","gene-321850","gene-321859","gene-321868","gene-321913","gene-321925","gene-321991","gene-322044","gene-322092","gene-322136","gene-322154","gene-322160","gene-342132","gene-342138","gene-342144","gene-342150","gene-342165","gene-342177","gene-342183","gene-342189","gene-342198","gene-342204","gene-342210","gene-342216","gene-342222","gene-342228","gene-342240","gene-342246","gene-342252","gene-342261","gene-342270","gene-342279","gene-342285","gene-342291","gene-342297","gene-342303","gene-342309","gene-342315","gene-342321","gene-342330","gene-342339","gene-342348","gene-342354","gene-342360","gene-342366","gene-342372","gene-342378","gene-342384","gene-342390","gene-342399","gene-342408","gene-342417","gene-342423","gene-342429","gene-342438","gene-342447","gene-342456","gene-342465","gene-342471","gene-342477","gene-342483","gene-342489","gene-342495","gene-342501","gene-342510","gene-342519","gene-342525","gene-342531","gene-342540","gene-342546","gene-342552","gene-342561","gene-342567","gene-342573","gene-342579","gene-342594","gene-342639","gene-342660","gene-342667","gene-342682","gene-342688","gene-342694","gene-342718","gene-342950","gene-342962","gene-342971","gene-342977","gene-342983","gene-343084","gene-343093","gene-343099","gene-343117","gene-343123","gene-343165","gene-343194","gene-343203","gene-343215","gene-343242","gene-343369","gene-343376","gene-343453","gene-343500","gene-343547","gene-343580","gene-343648","gene-343660","gene-343684","gene-343696","gene-343708","gene-343717","gene-343726","gene-343747","gene-343838","gene-343924","gene-343933","gene-343951","gene-343971","gene-343989","gene-344019","gene-344052","gene-344096","gene-344111","gene-344117","gene-344147","gene-344156","gene-344189","gene-344201","gene-344210","gene-344537","gene-344549","gene-344555","gene-344564","gene-344945","gene-344978","gene-345019","gene-345031","gene-345061","gene-345067","gene-345073","gene-345094","gene-345100","gene-345115","gene-345129","gene-345135","gene-345162","gene-345180","gene-345195","gene-345222","gene-345246","gene-345258","gene-345273","gene-345288","gene-345306","gene-345372","gene-345392","gene-345419","gene-345515","gene-345524","gene-345545","gene-345557","gene-367203","gene-367224","gene-367230","gene-367236","gene-367251","gene-367267","gene-367277","gene-367351","gene-367360","gene-367366","gene-367381","gene-367393","gene-367399","gene-367408","gene-367414","gene-367420","gene-367427","gene-367433","gene-367439","gene-367448","gene-367454","gene-367460","gene-367469","gene-367478","gene-367484","gene-367490","gene-367502","gene-367517","gene-381161","gene-381182","gene-381188","gene-381194","gene-381200","gene-381209","gene-381218","gene-381237","gene-381246","gene-381259","gene-381280","gene-381292","gene-381304","gene-381310","gene-381322","gene-381331","gene-381337","gene-381349","gene-381355","gene-381361","gene-381370","gene-381376","gene-381394","gene-381436","gene-381442","gene-381451","gene-381491","gene-381512","gene-381547","gene-381574","gene-381597","gene-381624","gene-381651","gene-381698","gene-381757","gene-381766","gene-381772","gene-381799","gene-381805","gene-381988","gene-381997","gene-382021","gene-382055","gene-382088","gene-382095","gene-382158","gene-382176","gene-382294","gene-382300","gene-382309","gene-382321","gene-382327","gene-382333","gene-382345","gene-382351","gene-382357","gene-382363","gene-382369","gene-382375","gene-382384","gene-382396","gene-382402","gene-382408","gene-382414","gene-382432","gene-382438","gene-382444","gene-382455","gene-382464","gene-382473","gene-382479","gene-382485","gene-382491","gene-382500","gene-382506","gene-382512","gene-382518","gene-382524","gene-382530","gene-382536","gene-382547","gene-382559","gene-382571","gene-382577","gene-382583","gene-382589","gene-382601","gene-382607","gene-382613","gene-382619","gene-382625","gene-386046","gene-386052","gene-386058","gene-386118","gene-386151","gene-386178","gene-386234","gene-386270","gene-386276","gene-386303","gene-386309","gene-386321","gene-386336","gene-386375","gene-386470","gene-386717","gene-386821","gene-386830","gene-386836","gene-386842","gene-386848","gene-386933","gene-386954","gene-386960","gene-386969","gene-386975","gene-386981","gene-386987","gene-386993","gene-387002","gene-387081","gene-387251","gene-387272","gene-387443","gene-387497","gene-387749","gene-387992","gene-388013","gene-388076","gene-388126","gene-388138","gene-388159","gene-388171","gene-388183","gene-388204","gene-388216","gene-388234","gene-388249","gene-388261","gene-388296","gene-388322","gene-388340","gene-388416","gene-388422","gene-388434","gene-388479","gene-388488","gene-388556","gene-388577","gene-388598","gene-419623","gene-419632")
y <- y[rownames(y$counts) %in% x_genes, , keep.lib.sizes=FALSE]
# remove samples that have 0 expression in all
y <- y[, colSums(y$counts) > 0]
y$samples$lib.size <- colSums(y$counts)
dim(y)
print(rownames(y$counts))
```

# basic DE with only one sex

I will partition the data to avoid having to handle a contrast with a three-way interaction. I will do first males and females separately to 
look at line and age effects, then lines spearately to look at age and sex effects, and then days separately to look at sex and line effects.

Since the females are supposed to be the same between both lines I hope to see minimal line difference when females are compared but that seems
to not always be the case.

## Males and Females separately

The analysis can be run separately for males and females, if use_males=True then it's only males, and if False then it's only the females

```{r} 
use_males<-FALSE
```

```{r} 
if (use_males){
## filter count data to only include males 
counts_M <- count_data[,grepl("-M_", colnames(count_data))]
}else{
## filter count data to only include females 
counts_M <- count_data[,grepl("-F_", colnames(count_data))]
}

meta_data <- DE_metadata_all[which((DE_metadata_all$ID %in% names(counts_M))==TRUE), ]
## adjust metadata
meta_data <- meta_data %>% remove_rownames %>% column_to_rownames(var="ID")
# reorder so it's in the same order as the count data columns
meta_data <- meta_data[names(counts_M),]
# check assumptions, should be TRUE
all(colnames(counts_M) %in% rownames(meta_data))

# make sample groups that reflect line and day
groups <- factor(paste(meta_data$line, meta_data$day, sep="_"))
# read data into data structure
y <- DGEList(counts=counts_M, group=groups)

## filter out low count genes
# filter by only line group and not day, since we aren't sure how the expression changes between days and we don't want to remove genes that are only rarely expressed
keep <- filterByExpr(y, group=meta_data$line, min.count=5, min.total.count=15)
  y <- y[keep, , keep.lib.sizes=FALSE]
  y$samples$lib.size <- colSums(y$counts) # recompute library sizes
  
# normalize by library sizes
y <- calcNormFactors(y)

# estimate dispersion
MM <- model.matrix(~0 + groups, data = y$samples)
colnames(MM) <- c("SL1_14", "SL1_16", "SL1_18", "SL3_14",  "SL3_16", "SL3_18")
rownames(MM) <- colnames(y)

y <- estimateDisp(y, MM)
# plotBCV(y)

```

Plot MDS for the sex-separated samples
```{r} 
if (use_males){
short_rownames <- c("1-14","1-14","1-14","1-14","1-14","1-14","1-14","1-14","1-14","1-14","1-16","1-16","1-16","1-16","1-16","1-16","1-16","1-16","1-16","1-16","1-16","1-18","1-18","1-18","1-18","1-18","1-18","1-18","3-14","3-14","3-14","3-14","3-14","3-14","3-14","3-14","3-14","3-16","3-16","3-16","3-16","3-16","3-16","3-16","3-18","3-18","3-18","3-18","3-18","3-18","3-18","3-18","3-18","3-18","3-18","3-18")
png("/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_figures/MDS_males_only.png", width = 7, height = 6, units = "in", res = 300)
plotMDS(y,  labels = short_rownames, col=as.numeric(y$samples$group))
title("MDS clustering of male samples")
dev.off()
}else{
short_rownames <- c("1_14","1_14","1_14","1_16","1_16","1_16","1_18","1_18","1_18","1_18","1_18","3_14","3_14","3_14","3_14","3_14","3_14","3_16","3_16","3_16","3_16","3_16","3_16","3_18","3_18","3_18")
png("/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_figures/MDS_females_only.png", width = 7, height = 6, units = "in", res = 300)
plotMDS(y,  labels = short_rownames, col=as.numeric(y$samples$group))
title("MDS clustering of female samples")
dev.off() 
}
```

### Differential expression just one sex (determined in the above section)

Contrasts are within day between line and pairwise comparisons within lines between days, as well as interactions of line-by-day.

```{r}
my.contrasts <- makeContrasts(
day14 =	SL1_14 - SL3_14, 
day16 =	SL1_16 - SL3_16, 
day18 = SL1_18 - SL3_18,
line_1_18_14 = SL1_18 - SL1_14,
line_1_14_16 = SL1_14 - SL1_16,
line_1_18_16 = SL1_18 - SL1_16,
line_3_18_14 = SL3_18 - SL3_14,
line_3_18_16 = SL3_18 - SL3_16,
line_3_14_16 = SL3_14 - SL3_16,
line_1_3_14_16 = (SL1_14 - SL3_14) - (SL1_16 - SL3_16),
line_1_3_18_16 = (SL1_18 - SL3_18) - (SL1_16 - SL3_16),
line_1_3_18_14 = (SL1_18 - SL3_18) - (SL1_14 - SL3_14),
levels=MM)

contrast_analysis <- function(y,MM,contrast, table_name, lfc_min){
  
  fit <- glmQLFit(y, MM)
  qlf <- glmQLFTest(fit, contrast=contrast)
  de_qlf <- decideTestsDGE(qlf, p.value=0.05, lfc=lfc_min)
  print(contrast)
  print(summary(de_qlf))
  
  print("--------------------------------------------")
  
  ## make output table
  Tags <- topTags(qlf, n=Inf, p.value=1)						
  Tags <- Tags$table
  Tags$Gene <- rownames(Tags)
  write.table(Tags, file=table_name, sep="\t")	
  
  output <- list(qlf=qlf,Tags=Tags)
  return(output)
}

```

line contrasts for each day

```{r}
if (use_males){
d14_table = sprintf("%sDE_genes_M_1-3_day14.txt", outdir)
}else{
d14_table = sprintf("%sDE_genes_F_1-3_day14.txt", outdir)
}
out_14 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"day14"],table_name=d14_table,lfc_min = 0)

if (use_males){
d16_table = sprintf("%sDE_genes_M_1-3_day16.txt", outdir)
}else{
d16_table = sprintf("%sDE_genes_F_1-3_day16.txt", outdir)
}
out_16 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"day16"],table_name=d16_table,lfc_min = 0)

if (use_males){
d18_table = sprintf("%sDE_genes_M_1-3_day18.txt", outdir)
}else{
d18_table = sprintf("%sDE_genes_F_1-3_day18.txt", outdir)
}
out_18 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"day18"],table_name=d18_table,lfc_min = 0)
```

Pairwise day contrasts within each line

```{r}
if (use_males){
line3_table = sprintf("%sDE_genes_M_SL3_18_14.txt", outdir)
}else{
line3_table = sprintf("%sDE_genes_F_SL3_18_14.txt", outdir)
}
out_SL3_18_14 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"line_3_18_14"],table_name=line3_table,lfc_min = 0)


if (use_males){
line3_table = sprintf("%sDE_genes_M_SL3_18_16.txt", outdir)
}else{
line3_table = sprintf("%sDE_genes_F_SL3_18_16.txt", outdir)
}
out_SL3_18_16 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"line_3_18_16"],table_name=line3_table,lfc_min = 0)

if (use_males){
line3_table = sprintf("%sDE_genes_M_SL3_14_16.txt", outdir)
}else{
line3_table = sprintf("%sDE_genes_F_SL3_14_16.txt", outdir)
}
out_SL3_14_16 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"line_3_14_16"],table_name=line3_table,lfc_min = 0)


if (use_males){
line1_table = sprintf("%sDE_genes_M_SL1_14_16.txt", outdir)
}else{
line1_table = sprintf("%sDE_genes_F_SL1_14_16.txt", outdir)
}
out_SL1_14_16 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"line_1_14_16"],table_name=line1_table,lfc_min = 0)


if (use_males){
line1_table = sprintf("%sDE_genes_M_SL1_18_14.txt", outdir)
}else{
line1_table = sprintf("%sDE_genes_F_SL1_18_14.txt", outdir)
}
out_SL1_18_14 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"line_1_18_14"],table_name=line1_table,lfc_min = 0)


if (use_males){
line1_table = sprintf("%sDE_genes_M_SL1_18_16.txt", outdir)
}else{
line1_table = sprintf("%sDE_genes_F_SL1_18_16.txt", outdir)
}
out_SL1_18_16 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"line_1_18_16"],table_name=line1_table,lfc_min = 0)

```

Interaction line by age

```{r}
if (use_males){
line1_3_table = sprintf("%sDE_genes_M_SL_1_3_18_14.txt", outdir)
}else{
line1_3_table = sprintf("%sDE_genes_F_SL_1_3_18_14.txt", outdir)
}
out_SL1_18_16 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"line_1_3_18_14"],table_name=line1_3_table,lfc_min = 0)

if (use_males){
line1_3_table = sprintf("%sDE_genes_M_SL_1_3_14_16.txt", outdir)
}else{
line1_3_table = sprintf("%sDE_genes_F_SL_1_3_14_16.txt", outdir)
}
out_SL1_18_16 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"line_1_3_14_16"],table_name=line1_3_table,lfc_min = 0)

if (use_males){
line1_3_table = sprintf("%sDE_genes_M_SL_1_3_18_16.txt", outdir)
}else{
line1_3_table = sprintf("%sDE_genes_F_SL_1_3_18_16.txt", outdir)
}
out_SL1_18_16 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"line_1_3_18_16"],table_name=line1_3_table,lfc_min = 0)

```

## Developmental sex differences (split by line)

I will see if there are sex differences in the different developmental stages within each line. Specify which line first.

```{r}
use_SL1 <- FALSE
```

```{r} 
if (use_SL1){
## filter count data to only include males 
counts_M <- count_data[,grepl("-1-", colnames(count_data))]
}else{
## filter count data to only include females 
counts_M <- count_data[,grepl("-3-", colnames(count_data))]
}

meta_data <- DE_metadata_all[which((DE_metadata_all$ID %in% names(counts_M))==TRUE), ]
## adjust metadata
meta_data <- meta_data %>% remove_rownames %>% column_to_rownames(var="ID")
# reorder so it's in the same order as the count data columns
meta_data <- meta_data[names(counts_M),]
# check assumptions, should be TRUE
all(colnames(counts_M) %in% rownames(meta_data))

# make sample groups that reflect line and day
groups <- factor(paste(meta_data$sex, meta_data$day, sep="_"))
# read data into data structure
y <- DGEList(counts=counts_M, group=groups)

## filter out low count genes
keep <- filterByExpr(y, group=meta_data$sex, min.count=5, min.total.count=15)
  y <- y[keep, , keep.lib.sizes=FALSE]
  y$samples$lib.size <- colSums(y$counts) # recompute library sizes
  
# normalize by library sizes
y <- calcNormFactors(y)

# estimate dispersion
MM <- model.matrix(~0 + groups, data = y$samples)
colnames(MM) <- c("F_14", "F_16", "F_18", "M_14",  "M_16", "M_18")
rownames(MM) <- colnames(y)

y <- estimateDisp(y, MM)
# plotBCV(y)
```

```{r} 
if (use_SL1){
short_rownames <- c("M_14","F_14","M_14","M_14","F_14","M_14","M_14","M_14","M_14","F_14","M_14","M_14","M_14","M_16","M_16","M_16","F_16","M_16","F_16","M_16","F_16","M_16","M_16","M_16","M_16","M_16","M_16","M_18","M_18","F_18","F_18","M_18","M_18","M_18","M_18","M_18","F_18","F_18","F_18")
png("/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_figures/MDS_SL1_only.png", width = 7, height = 6, units = "in", res = 300)
plotMDS(y,  labels = short_rownames, col=as.numeric(y$samples$group))
title("MDS clustering of SL1 samples")
dev.off()
}else{
short_rownames <- c("M_14","M_14","F_14","M_14","M_14","F_14","M_14","F_14","M_14","M_14","F_14","M_14","F_14","F_14","M_14","F_16","M_16","F_16","M_16","M_16","F_16","M_16","F_16","F_16","M_16","M_16","M_16","F_16","F_18","M_18","M_18","M_18","M_18","M_18","M_18","M_18","F_18","M_18","F_18","M_18","M_18","M_18","M_18")
png("/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_figures/MDS_SL3_only.png", width = 7, height = 6, units = "in", res = 300)
plotMDS(y,  labels = short_rownames, col=as.numeric(y$samples$group))
title("MDS clustering of SL3 samples")
dev.off() 
}
```

I will use the same contrast analysis function as above and just give new constants

```{r}
my.contrasts <- makeContrasts(
day14 =	F_14 - M_14, 
day16 =	F_16 - M_16, 
day18 = F_18 - M_18,
day14_16 = (F_14 - M_14) - (F_16 - M_16),
day18_14 = (F_18 - M_18) - (F_14 - M_14),
day18_16 = (F_18 - M_18) - (F_16 - M_16),
levels=MM)
```

Within-day sex bias 

```{r}
if (use_SL1){
d14_table = sprintf("%sDE_genes_SL1_day14_F-M.txt", outdir)
}else{
d14_table = sprintf("%sDE_genes_SL3_day14_F-M.txt", outdir)
}
out_14 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"day14"],table_name=d14_table,lfc_min = 0)

if (use_SL1){
d16_table = sprintf("%sDE_genes_SL1_day16_F-M.txt", outdir)
}else{
d16_table = sprintf("%sDE_genes_SL3_day16_F-M.txt", outdir)
}
out_16 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"day16"],table_name=d16_table,lfc_min = 0)

if (use_SL1){
d18_table = sprintf("%sDE_genes_SL1_day18_F-M.txt", outdir)
}else{
d18_table = sprintf("%sDE_genes_SL3_day18_F-M.txt", outdir)
}
out_18 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"day18"],table_name=d18_table,lfc_min = 0)
```

sex-bias by day-comparison interaction

```{r}
if (use_SL1){
d14_16_table = sprintf("%sDE_genes_SL1_day14_16_F-M.txt", outdir)
}else{
d14_16_table = sprintf("%sDE_genes_SL3_day14_16_F-M.txt", outdir)
}
out_14 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"day14_16"],table_name=d14_16_table,lfc_min = 0)

if (use_SL1){
d18_14_table = sprintf("%sDE_genes_SL1_day18_14_F-M.txt", outdir)
}else{
d18_14_table = sprintf("%sDE_genes_SL3_day18_14_F-M.txt", outdir)
}
out_16 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"day18_14"],table_name=d18_14_table,lfc_min = 0)

if (use_SL1){
d18_16_table = sprintf("%sDE_genes_SL1_day18_16_F-M.txt", outdir)
}else{
d18_16_table = sprintf("%sDE_genes_SL3_day18_16_F-M.txt", outdir)
}
out_18 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"day18_16"],table_name=d18_16_table,lfc_min = 0)
```

## Line sex difference (split by day)

Look at the sex-bias and line stuff by splitting the data by day

```{r}
day <- "14" # "14" or "16" or "18"
```

```{r} 
day1 <- paste("-1-", day, sep="") # do it like this to not match the sample number, only the actual day
day3 <- paste("-3-", day, sep="")
counts_M <- count_data[, grepl(day1, colnames(count_data)) | grepl(day3, colnames(count_data))]

meta_data <- DE_metadata_all[which((DE_metadata_all$ID %in% names(counts_M))==TRUE), ]
## adjust metadata
meta_data <- meta_data %>% remove_rownames %>% column_to_rownames(var="ID")
# reorder so it's in the same order as the count data columns
meta_data <- meta_data[names(counts_M),]
# check assumptions, should be TRUE
all(colnames(counts_M) %in% rownames(meta_data))

# make sample groups that reflect line and day
groups <- factor(paste(meta_data$sex, meta_data$line, sep="_"))
# read data into data structure
y <- DGEList(counts=counts_M, group=groups)

## filter out low count genes
keep <- filterByExpr(y, group=meta_data$sex, min.count=5, min.total.count=15)
  y <- y[keep, , keep.lib.sizes=FALSE]
  y$samples$lib.size <- colSums(y$counts) # recompute library sizes
  
# normalize by library sizes
y <- calcNormFactors(y)

# estimate dispersion
MM <- model.matrix(~0 + groups, data = y$samples)
colnames(MM) <- c("F_1", "F_3", "M_1", "M_3")
rownames(MM) <- colnames(y)

y <- estimateDisp(y, MM)
# plotBCV(y)
```

I don't do the mds plot for these ones because it's weird with the sample names. maybe later at some point. Define contrasts next

```{r}
my.contrasts <- makeContrasts(
line1 =	F_1 - M_1,
line3 =	F_3 - M_3,
females = F_1 - F_3,
males = M_1 - M_3,
sex_line = (F_1 - M_1) - (F_3 - M_3),
levels=MM)
```

```{r}
SL1_table <- sprintf("/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_genes_day%s_SL1_F-M.txt", day)
out_SL1 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"line1"],table_name=SL1_table,lfc_min = 0)

SL3_table <- sprintf("/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_genes_day%s_SL3_F-M.txt", day)
out_SL3 <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"line3"],table_name=SL3_table,lfc_min = 0)

females_table <- sprintf("/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_genes_day%s_F_1-3.txt", day)
out_f <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"females"],table_name=females_table,lfc_min = 0)

males_table <- sprintf("/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_genes_day%s_M_1-3.txt", day)
out_m <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"males"],table_name=males_table,lfc_min = 0)

interaction_table <- sprintf("/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_genes_day%s_F-M_by_1-3.txt", day)
out_m <- contrast_analysis(y=y,MM=MM,contrast=my.contrasts[,"males"],table_name=interaction_table,lfc_min = 0)
```