---
title: "DE_analysis"
author: "Milena Trabert"
date: "2026-03-27"
output: html_document
---

# GO term enrichment on the significant gene sets 

I have lists of significantly DE genes in all contrasts. This is the same as the GO enrichment for chapter3.
Detailed documentation is here: https://rdrr.io/bioc/topGO/f/inst/doc/topGO.pdf

## Install libraries

```{r}
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install()
BiocManager::install("topGO")
BiocManager::install("GO.db")
BiocManager::install("ViSEAGO")
```

```{r}
library(topGO)
library(plyr)
library(ViSEAGO)
library(stringr)
```

## read input files

annotation and background gene set

```{r}
annotation_path="/Users/miltr339/work/c_maculatus/C_mac_eggnog_diamond.emapper.annotations_geneIDs" # created from the eggnogmapper output with /Users/miltr339/work/PhD_code/PhD_chapter4/src/make_annot_with_geneIDs.py

make_annotation_df_from_eggnog <- function(annotation_path){

    annot_header = c("query","seed_ortholog","evalue","score","eggNOG_OGs","max_annot_lvl","COG_category","Description","Preferred_name","GOs","EC","KEGG_ko","KEGG_Pathway","KEGG_Module","KEGG_Reaction","KEGG_rclass","BRITE","KEGG_TC","CAZy","BiGG_Reaction","PFAMs")
    annotation = read.csv(annotation_path, sep="\t", comment.char = "#", header=FALSE, col.names = annot_header)
    background_geneIDs <- annotation$query
    # prepare for input
    valid <- !is.na(annotation$GOs) & annotation$GOs != "-"
    annotation_df <- strsplit(annotation$GOs[valid], ",")
    names(annotation_df) <- annotation$query[valid]

    return(annotation_df)
}

annotation_df <- make_annotation_df_from_eggnog(annotation_path)

```

lists of significant geneIDs for all contrasts

```{r}
indir <- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/sig_DE_genes_lists"
# sex separated
## females
sex_sep_F <- c(sprintf("%s/sig_DE_list_F_1-3_day14.txt",indir),sprintf("%s/sig_DE_list_F_1-3_day16.txt",indir),sprintf("%s/sig_DE_list_F_1-3_day18.txt",indir),sprintf("%s/sig_DE_list_F_SL1.txt",indir),sprintf("%s/sig_DE_list_F_SL1_14_16.txt",indir),sprintf("%s/sig_DE_list_F_SL1_18_14.txt",indir),sprintf("%s/sig_DE_list_F_SL1_18_16.txt",indir),sprintf("%s/sig_DE_list_F_SL3.txt",indir),sprintf("%s/sig_DE_list_F_SL3_14_16.txt",indir),sprintf("%s/sig_DE_list_F_SL3_18_14.txt",indir),sprintf("%s/sig_DE_list_F_SL3_18_16.txt",indir),sprintf("%s/sig_DE_list_F_SL_1_3_14_16.txt",indir),sprintf("%s/sig_DE_list_F_SL_1_3_18_14.txt",indir),sprintf("%s/sig_DE_list_F_SL_1_3_18_16.txt",indir))
## males
sex_sep_M <- c(sprintf("%s/sig_DE_list_M_1-3_day14.txt",indir),sprintf("%s/sig_DE_list_M_1-3_day16.txt",indir),sprintf("%s/sig_DE_list_M_1-3_day18.txt",indir),sprintf("%s/sig_DE_list_M_SL1.txt",indir),sprintf("%s/sig_DE_list_M_SL1_14_16.txt",indir),sprintf("%s/sig_DE_list_M_SL1_18_14.txt",indir),sprintf("%s/sig_DE_list_M_SL1_18_16.txt",indir),sprintf("%s/sig_DE_list_M_SL3.txt",indir),sprintf("%s/sig_DE_list_M_SL3_14_16.txt",indir),sprintf("%s/sig_DE_list_M_SL3_18_14.txt",indir),sprintf("%s/sig_DE_list_M_SL3_18_16.txt",indir),sprintf("%s/sig_DE_list_M_SL_1_3_14_16.txt",indir),sprintf("%s/sig_DE_list_M_SL_1_3_18_14.txt",indir),sprintf("%s/sig_DE_list_M_SL_1_3_18_16.txt",indir))

# line separated
## SL1
line_sep_SL1 <- c(sprintf("%s/sig_DE_list_SL1_day14_16_F-M.txt",indir),sprintf("%s/sig_DE_list_SL1_day14_F-M.txt",indir),sprintf("%s/sig_DE_list_SL1_day16_F-M.txt",indir),sprintf("%s/sig_DE_list_SL1_day18_14_F-M.txt",indir),sprintf("%s/sig_DE_list_SL1_day18_16_F-M.txt",indir),sprintf("%s/sig_DE_list_SL1_day18_F-M.txt",indir))
## SL3
line_sep_SL3 <- c(sprintf("%s/sig_DE_list_SL3_day14_16_F-M.txt",indir),sprintf("%s/sig_DE_list_SL3_day14_F-M.txt",indir),sprintf("%s/sig_DE_list_SL3_day16_F-M.txt",indir),sprintf("%s/sig_DE_list_SL3_day18_14_F-M.txt",indir),sprintf("%s/sig_DE_list_SL3_day18_16_F-M.txt",indir),sprintf("%s/sig_DE_list_SL3_day18_F-M.txt",indir))

# day separated
## day 14
day_sep_d14 <- c(sprintf("%s/sig_DE_list_day14_F-M_by_1-3.txt",indir),sprintf("%s/sig_DE_list_day14_F_1-3.txt",indir),sprintf("%s/sig_DE_list_day14_M_1-3.txt",indir),sprintf("%s/sig_DE_list_day14_SL1_F-M.txt",indir),sprintf("%s/sig_DE_list_day14_SL3_F-M.txt",indir))
## day 16
day_sep_d16 <- c(sprintf("%s/sig_DE_list_day16_F-M_by_1-3.txt",indir),sprintf("%s/sig_DE_list_day16_F_1-3.txt",indir),sprintf("%s/sig_DE_list_day16_M_1-3.txt",indir),sprintf("%s/sig_DE_list_day16_SL1_F-M.txt",indir),sprintf("%s/sig_DE_list_day16_SL3_F-M.txt",indir))
## day 18
day_sep_d18 <- c(sprintf("%s/sig_DE_list_day18_F-M_by_1-3.txt",indir),sprintf("%s/sig_DE_list_day18_F_1-3.txt",indir),sprintf("%s/sig_DE_list_day18_M_1-3.txt",indir),sprintf("%s/sig_DE_list_day18_SL1_F-M.txt",indir),sprintf("%s/sig_DE_list_day18_SL3_F-M.txt",indir))
```

## run enrichment

make GO enrichment object for each contrast
```{r}

GO_enrichment_analysis <- function(sig_infile, annotation_df){
  # prepare input data
  sig_df <- read.csv(sig_infile, sep=",", header=TRUE)
  geneList <- factor(as.integer(sig_df$sig_DE))
  names(geneList) <- sig_df$geneID
  ## downsample annotation to only include genes in the current test
  counts <- count(geneList)
  sigs_num <- length(geneList[geneList==1])

  ## only run if there are at least 10 significantly differentially expressed genes!
  if (sigs_num > 9){
    gene2GO <- annotation_df[names(annotation_df) %in% names(geneList)]
    ## create topGO object
    # ontology = BP means that it's all biological process level GOs
    GOdata <- new("topGOdata", ontology = "BP", allGenes = geneList, annot = annFUN.gene2GO, gene2GO = gene2GO)
    # run enrichment test
    result <- runTest(GOdata, algorithm = "elim", statistic = "fisher")
    ## about the "elim" algorithm from the documentation: "Here one computes the p-value of a GO term conditioned on the neighbouring terms. The tests are therefore not independent and the multiple testing theory does not directly apply. We like to interpret the p-values returned by these methods as corrected or not affected by multiple testing."
    allRes <- GenTable(GOdata,classicFisher = result,orderBy = "classicFisher",topNodes = 1000)
    # write to outfile
    outfile_name <- str_replace(sig_infile, "sig_DE_list_", "GO_enrichment_")
    write.csv(allRes, outfile_name, row.names=TRUE)
    return(allRes)
  }
}
```

loop through all lists
```{r}

for (sig_list in sex_sep_F){GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df)}
for (sig_list in sex_sep_M){GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df)}

for (sig_list in line_sep_SL1){GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df)}
for (sig_list in line_sep_SL3){GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df)}

for (sig_list in day_sep_d14){GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df)}
for (sig_list in day_sep_d16){GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df)}
for (sig_list in day_sep_d18){GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df)}

```

list for genes that are sig line-biased in males in both day14 and day16
```{r}
sig_list = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/sig_DE_genes_lists/sig_DE_list_day_sep_M1-M3_day14and16.txt"
GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df)
```