---
title: "DE_analysis"
author: "Milena Trabert"
date: "2026-03-27"
output: html_document
---

# GO term enrichment on the significant gene sets 

I have lists of significantly DE genes in all contrasts. This is the same as the GO enrichment for chapter3

## Install libraries

```{r}
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install()
BiocManager::install("topGO")
BiocManager::install("GO.db")
BiocManager::install("ViSEAGO")
```

```{r}
library(topGO)
library(plyr)
library(ViSEAGO)
```

## read input files

annotation and background gene set

```{r}
annotation_path="/Users/miltr339/work/c_maculatus/C_mac_eggnog_diamond.emapper.annotations_geneIDs" # created from the eggnogmapper output with /Users/miltr339/work/PhD_code/PhD_chapter4/src/make_annot_with_geneIDs.py
annot_header = c("query","seed_ortholog","evalue","score","eggNOG_OGs","max_annot_lvl","COG_category","Description","Preferred_name","GOs","EC","KEGG_ko","KEGG_Pathway","KEGG_Module","KEGG_Reaction","KEGG_rclass","BRITE","KEGG_TC","CAZy","BiGG_Reaction","PFAMs")
annotation = read.csv(annotation_path, sep="\t", comment.char = "#", header=FALSE, col.names = annot_header)
```

lists of significant geneIDs for all contrasts

```{r}
indir <- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/sig_DE_genes_lists"
# sex separated
## females
sig_list <- sprintf("%s/sig_DE_list_F_1-3_day14.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_F_1-3_day16.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_F_1-3_day18.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_F_SL1.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_F_SL1_14_16.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_F_SL1_18_14.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_F_SL1_18_16.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_F_SL3.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_F_SL3_14_16.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_F_SL3_18_14.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_F_SL3_18_16.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_F_SL_1_3_14_16.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_F_SL_1_3_18_14.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_F_SL_1_3_18_16.txt",indir)
## males
sig_list <- sprintf("%s/sig_DE_list_M_1-3_day14.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_M_1-3_day16.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_M_1-3_day18.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_M_SL1.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_M_SL1_14_16.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_M_SL1_18_14.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_M_SL1_18_16.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_M_SL3.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_M_SL3_14_16.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_M_SL3_18_14.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_M_SL3_18_16.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_M_SL_1_3_14_16.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_M_SL_1_3_18_14.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_M_SL_1_3_18_16.txt",indir)

# line separated
## SL1
sig_list <- sprintf("%s/sig_DE_list_SL1_day14_16_F-M.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_SL1_day14_F-M.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_SL1_day16_F-M.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_SL1_day18_14_F-M.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_SL1_day18_16_F-M.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_SL1_day18_F-M.txt",indir)
## SL3
sig_list <- sprintf("%s/sig_DE_list_SL3_day14_16_F-M.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_SL3_day14_F-M.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_SL3_day16_F-M.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_SL3_day18_14_F-M.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_SL3_day18_16_F-M.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_SL3_day18_F-M.txt",indir)

# day separated
## day14
sig_list <- sprintf("%s/sig_DE_list_day14_F-M_by_1-3.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_day14_F_1-3.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_day14_M_1-3.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_day14_SL1_F-M.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_day14_SL3_F-M.txt",indir)
## day 16
sig_list <- sprintf("%s/sig_DE_list_day16_F-M_by_1-3.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_day16_F_1-3.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_day16_M_1-3.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_day16_SL1_F-M.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_day16_SL3_F-M.txt",indir)
## day 18
sig_list <- sprintf("%s/sig_DE_list_day18_F-M_by_1-3.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_day18_F_1-3.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_day18_M_1-3.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_day18_SL1_F-M.txt",indir)
sig_list <- sprintf("%s/sig_DE_list_day18_SL3_F-M.txt",indir)
```

make into proper genelist that topGO can use
```{r}
make_gene_list <- function(input_df){
  geneList <- factor(as.integer(input_df$pos_sel))
  names(geneList) <- input_df$geneID
  return(geneList)
}

enrichment_list <- make_gene_list(sig_list)
```

### Make into topGO data structure: 
start with making the enrichment geneIDs lists into the required named vectors 
(since I have binary categories and not p-values as expected, the data frames have values that are 1 if the gene is positively selected and 0 otherwise)
```{r}
sig_list = strsplit(readLines(sig_infile), ",")[[1]]

```