---
title: "DE_analysis"
author: "Milena Trabert"
date: "2026-03-27"
output: html_document
---

# GO term enrichment on the significant gene sets 

I have lists of significantly DE genes in all contrasts. This is the same as the GO enrichment for chapter3.
Detailed documentation is here: https://rdrr.io/bioc/topGO/f/inst/doc/topGO.pdf

## Install libraries

```{r}
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")

BiocManager::install("topGO")
BiocManager::install("GO.db")
```
```{r}

### simplifyEnrichment package and dependencies:

# There is an error in GlobalOptions where it imports findMatches from a wrong namespace that i fixed like this in the source code:
# 1. download from source: 
#     curl -O https://cloud.r-project.org/src/contrib/GlobalOptions_0.1.4.tar.gz
#     tar -xzf GlobalOptions_0.1.4.tar.gz
# 2. find the wrong import:
#     grep -n "findMatches" GlobalOptions/NAMESPACE
# 3. remove the wrong line:
#     sed -i '' '/importFrom(utils,findMatches)/d' GlobalOptions/NAMESPACE
# 4. build:
#     R CMD build GlobalOptions
if(FALSE){
  install.packages("/Users/miltr339/work/software/GlobalOptions_0.1.4.tar.gz", repos = NULL, type = "source")
  library(GlobalOptions)
  install.packages("RcppParallel")
  library(RcppParallel)
  BiocManager::install("simplifyEnrichment")
}
```

```{r}
library(topGO)
library(plyr)
library(stringr)
library(GlobalOptions)
library(RcppParallel)
library(simplifyEnrichment)
```

## read input files

annotation and background gene set

```{r}
make_annotation_df_from_eggnog <- function(annotation_path){
    annot_header = c("query","seed_ortholog","evalue","score","eggNOG_OGs","max_annot_lvl","COG_category","Description","Preferred_name","GOs","EC","KEGG_ko","KEGG_Pathway","KEGG_Module","KEGG_Reaction","KEGG_rclass","BRITE","KEGG_TC","CAZy","BiGG_Reaction","PFAMs")
    annotation <- read.csv(annotation_path, sep="\t", comment.char = "#", header=FALSE, col.names = annot_header)
    background_geneIDs <- annotation$query
    # prepare for input
    valid <- !is.na(annotation$GOs) & annotation$GOs != "-"
    annotation_df <- strsplit(annotation$GOs[valid], ",")
    names(annotation_df) <- annotation$query[valid]

    return(annotation_df)
}

annotation_path="/Users/miltr339/work/c_maculatus/C_mac_eggnog_diamond.emapper.annotations_geneIDs" # created from the eggnogmapper output with /Users/miltr339/work/PhD_code/PhD_chapter4/src/make_annot_with_geneIDs.py

annotation_df <- make_annotation_df_from_eggnog(annotation_path)
```

lists of significant geneIDs for all contrasts

```{r}
indir <- "/Users/miltr339/work/PhD_code/PhD_chapter4/data/sig_DE_genes_lists"
# sex separated
## females
sex_sep_F <- c(sprintf("%s/sig_DE_list_F_1-3_day14.txt",indir),sprintf("%s/sig_DE_list_F_1-3_day16.txt",indir),sprintf("%s/sig_DE_list_F_1-3_day18.txt",indir),sprintf("%s/sig_DE_list_F_SL1.txt",indir),sprintf("%s/sig_DE_list_F_SL1_14_16.txt",indir),sprintf("%s/sig_DE_list_F_SL1_18_14.txt",indir),sprintf("%s/sig_DE_list_F_SL1_18_16.txt",indir),sprintf("%s/sig_DE_list_F_SL3.txt",indir),sprintf("%s/sig_DE_list_F_SL3_14_16.txt",indir),sprintf("%s/sig_DE_list_F_SL3_18_14.txt",indir),sprintf("%s/sig_DE_list_F_SL3_18_16.txt",indir),sprintf("%s/sig_DE_list_F_SL_1_3_14_16.txt",indir),sprintf("%s/sig_DE_list_F_SL_1_3_18_14.txt",indir),sprintf("%s/sig_DE_list_F_SL_1_3_18_16.txt",indir),sprintf("%s/only_F_no_line_sig_DE_list_day_14-16.txt",indir),sprintf("%s/only_F_no_line_sig_DE_list_day_16-18.txt",indir),sprintf("%s/only_F_no_line_sig_DE_list_day_14-18.txt",indir))
## males
sex_sep_M <- c(sprintf("%s/sig_DE_list_M_1-3_day14.txt",indir),sprintf("%s/sig_DE_list_M_1-3_day16.txt",indir),sprintf("%s/sig_DE_list_M_1-3_day18.txt",indir),sprintf("%s/sig_DE_list_M_SL1.txt",indir),sprintf("%s/sig_DE_list_M_SL1_14_16.txt",indir),sprintf("%s/sig_DE_list_M_SL1_18_14.txt",indir),sprintf("%s/sig_DE_list_M_SL1_18_16.txt",indir),sprintf("%s/sig_DE_list_M_SL3.txt",indir),sprintf("%s/sig_DE_list_M_SL3_14_16.txt",indir),sprintf("%s/sig_DE_list_M_SL3_18_14.txt",indir),sprintf("%s/sig_DE_list_M_SL3_18_16.txt",indir),sprintf("%s/sig_DE_list_M_SL_1_3_14_16.txt",indir),sprintf("%s/sig_DE_list_M_SL_1_3_18_14.txt",indir),sprintf("%s/sig_DE_list_M_SL_1_3_18_16.txt",indir),sprintf("%s/only_M_no_line_sig_DE_list_day_14-16.txt",indir),sprintf("%s/only_M_no_line_sig_DE_list_day_16-18.txt",indir),sprintf("%s/only_M_no_line_sig_DE_list_day_14-18.txt",indir))

# line separated
## SL1
line_sep_SL1 <- c(sprintf("%s/sig_DE_list_SL1_day14_16_F-M.txt",indir),sprintf("%s/sig_DE_list_SL1_day14_F-M.txt",indir),sprintf("%s/sig_DE_list_SL1_day16_F-M.txt",indir),sprintf("%s/sig_DE_list_SL1_day18_14_F-M.txt",indir),sprintf("%s/sig_DE_list_SL1_day18_16_F-M.txt",indir),sprintf("%s/sig_DE_list_SL1_day18_F-M.txt",indir))
## SL3
line_sep_SL3 <- c(sprintf("%s/sig_DE_list_SL3_day14_16_F-M.txt",indir),sprintf("%s/sig_DE_list_SL3_day14_F-M.txt",indir),sprintf("%s/sig_DE_list_SL3_day16_F-M.txt",indir),sprintf("%s/sig_DE_list_SL3_day18_14_F-M.txt",indir),sprintf("%s/sig_DE_list_SL3_day18_16_F-M.txt",indir),sprintf("%s/sig_DE_list_SL3_day18_F-M.txt",indir))

# day separated
## day 14
day_sep_d14 <- c(sprintf("%s/sig_DE_list_day14_SL1_F-M.txt",indir),sprintf("%s/sig_DE_list_day14_SL3_F-M.txt",indir))#,sprintf("%s/sig_DE_list_day14_M_1-3.txt",indir),sprintf("%s/sig_DE_list_day14_F-M_by_1-3.txt",indir),sprintf("%s/sig_DE_list_day14_F_1-3.txt",indir))
## day 16
day_sep_d16 <- c(sprintf("%s/sig_DE_list_day16_SL1_F-M.txt",indir),sprintf("%s/sig_DE_list_day16_SL3_F-M.txt",indir))#,sprintf("%s/sig_DE_list_day16_M_1-3.txt",indir),sprintf("%s/sig_DE_list_day16_F-M_by_1-3.txt",indir),sprintf("%s/sig_DE_list_day16_F_1-3.txt",indir))
## day 18
day_sep_d18 <- c(sprintf("%s/sig_DE_list_day18_SL1_F-M.txt",indir),sprintf("%s/sig_DE_list_day18_SL3_F-M.txt",indir))#,sprintf("%s/sig_DE_list_day18_M_1-3.txt",indir),sprintf("%s/sig_DE_list_day18_F-M_by_1-3.txt",indir),sprintf("%s/sig_DE_list_day18_F_1-3.txt",indir))

# no separation
## line ignored
line_ig_sex_bias <- c(sprintf("%s/full_dataset_line_ignored_day14_F-M.txt", indir),sprintf("%s/full_dataset_line_ignored_day16_F-M.txt", indir),sprintf("%s/full_dataset_line_ignored_day18_F-M.txt", indir)) 
line_ig_day_bias <- c(sprintf("%s/full_dataset_line_ignored_M_14_16.txt", indir),sprintf("%s/full_dataset_line_ignored_M_16_18.txt", indir)) 


```

## run enrichment

make GO enrichment object for each contrast
```{r}
GO_enrichment_analysis <- function(sig_infile, annotation_df, semantic_clustering=FALSE){
  # prepare input data
  sig_df <- read.csv(sig_infile, sep=",", header=TRUE)
  geneList <- factor(as.integer(sig_df$sig_DE))
  names(geneList) <- sig_df$geneID
  ## downsample annotation to only include genes in the current test
  counts <- count(geneList)
  sigs_num <- length(geneList[geneList==1])

  ## only run if there are at least 10 significantly differentially expressed genes! # nolint
  if (sigs_num > 9){
    gene2GO <- annotation_df[names(annotation_df) %in% names(geneList)]
    ## create topGO object
    # ontology = BP means that it's all biological process level GOs
    GOdata <- new("topGOdata", ontology = "BP", allGenes = geneList, annot = annFUN.gene2GO, gene2GO = gene2GO)
    # run enrichment test
    result <- runTest(GOdata, algorithm = "elim", statistic = "fisher")
    ## about the "elim" algorithm from the documentation: "Here one computes the p-value of a GO term conditioned on the neighbouring terms. The tests are therefore not independent and the multiple testing theory does not directly apply. We like to interpret the p-values returned by these methods as corrected or not affected by multiple testing."
    allRes <- GenTable(GOdata,classicFisher = result,orderBy = "classicFisher",topNodes = 1000)
    # make vector of sig. GO terms
    allRes$classicFisher <- as.numeric(gsub("< ", "", allRes$classicFisher))
    sigRes <- allRes[allRes$classicFisher < 0.05, ]
    go_ids <- sigRes$GO.ID
    # write to outfile
    outfile_name <- str_replace(sig_infile, "sig_DE_list_", "GO_enrichment_")
    outfile_name <- str_replace(sig_infile, "interaction", "interaction_GO_enrichment") 
    outfile_name <- str_replace(sig_infile, "ignore", "ignore_GO_enrichment") 
    outfile_name <- str_replace(outfile_name, ".txt", ".csv") 
    write.csv(sigRes, outfile_name, row.names=TRUE)

    ## make enrichment visualization
    # use simplifyEnrichment
    # this takes super long, so only do when it's relevant, not worth it for everything
    if(semantic_clustering){
      outfile_name <- str_replace(outfile_name, ".csv", ".png") 
      mat <- GO_similarity(go_ids, ont = "BP")  # biological process (BP) like above! needs to match
      # do the clustering, this takes quite a while
      png(outfile_name, width = 10, height = 8, units = "in", res = 300)
      df <- simplifyGO(mat, use_raster = FALSE)
      dev.off()
    }

    return(sigRes)
  }
}
```

loop through all lists
```{r}

for (sig_list in sex_sep_F){GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df)}
for (sig_list in sex_sep_M){GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df)}

for (sig_list in line_sep_SL1){GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df)}
for (sig_list in line_sep_SL3){GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df)}

for (sig_list in day_sep_d14){GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df, semantic_clustering=TRUE)}
for (sig_list in day_sep_d16){GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df, semantic_clustering=TRUE)}
for (sig_list in day_sep_d18){GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df, semantic_clustering=TRUE)}

for (sig_list in line_ig_sex_bias){GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df)}
for (sig_list in line_ig_day_bias){GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df, semantic_clustering=TRUE)}

```

list for genes that are sig line-biased in males in both day14 and day16
```{r}
sig_list = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/sig_DE_genes_lists/sig_DE_list_day_sep_M1-M3_day14and16.txt"
GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df)
```

list for genes that are sig in the day-sex interaction of the full dataset
```{r}
sig_list = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/sig_DE_genes_lists/full_dataset_line_ignored_day_sex_interaction.txt"
GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df)
```

list for genes that are sig. sex biased in the day separated dataset for multiple days
```{r}
sig_list_late = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/sig_DE_genes_lists/day_separated_all_lines_sex_bias_overlap_day1618_sigIDs.txt"
GO_enrichment_analysis(sig_infile = sig_list_late, annotation_df = annotation_df, semantic_clustering=TRUE)
sig_list_early = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/sig_DE_genes_lists/day_separated_all_lines_sex_bias_overlap_day14_sigIDs.txt"
GO_enrichment_analysis(sig_infile = sig_list_early, annotation_df = annotation_df, semantic_clustering=TRUE)
sig_list_all = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/sig_DE_genes_lists/day_separated_all_lines_sex_bias_overlap_all_days_sigIDs.txt"
GO_enrichment_analysis(sig_infile = sig_list_all, annotation_df = annotation_df, semantic_clustering=TRUE)
sig_list = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/sig_DE_genes_lists/day_separated_all_lines_sex_bias_overlap_all_excl_day14smallY_sigIDs.txt"
GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df, semantic_clustering=TRUE)
```

List of genes that are significant in the day-separated data for the line-by-sex interaction.
```{r}
sig_list = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/sig_DE_genes_lists/sig_DE_list_day14_F-M_by_1-3.txt"
GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df, semantic_clustering=TRUE)
sig_list = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/sig_DE_genes_lists/sig_DE_list_day16_F-M_by_1-3.txt"
GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df, semantic_clustering=TRUE)
# no day 18 because there's only 14 genes in the interaction
```

List of genes that are sex-biased in males and line-biased in SL3 (largeY) on day 16
This enrichment is based on the LFC comparison plots of the same name, and the background is all genes in the plot, so the genes that are sig. in at least one axis.
and the foreground is genes that are sig in both.
```{r}
sig_list = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/sig_DE_genes_lists/LFC_scatter_interaction_day16_sb_SL3_lb_M_shared_sig_DE_list.txt"
GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df, semantic_clustering=TRUE)
sig_list = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/sig_DE_genes_lists/LFC_scatter_interaction_day16_sb_SL1_lb_M_shared_sig_DE_list.txt"
GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df, semantic_clustering=TRUE)
sig_list = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/sig_DE_genes_lists/LFC_scatter_interaction_day14_sb_SL3_lb_M_shared_sig_DE_list.txt"
GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df, semantic_clustering=TRUE)
sig_list = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/sig_DE_genes_lists/LFC_scatter_interaction_day14_sb_SL1_lb_M_shared_sig_DE_list.txt"
GO_enrichment_analysis(sig_infile = sig_list, annotation_df = annotation_df, semantic_clustering=TRUE)
```


analyze GO term enrichment from upsetplot intersection lists
```{r}
semantic_clistering_from_table <- function(sig_infile){
  # prepare input data
  sig_df <- read.csv(sig_infile, sep="\t", header=FALSE)
  go_ids <- sig_df$V1
  sigs_num <- length(go_ids)

  ## only run if there are at least 10 significantly differentially expressed genes! # nolint
  if (sigs_num > 9){
    outfile_name <- str_replace(sig_infile, ".txt", ".png") 

    ## make enrichment visualization
    # use simplifyEnrichment
    mat <- GO_similarity(go_ids, ont = "BP")  # biological process (BP) like above! needs to match
    # do the clustering, this takes quite a while
    png(outfile_name, width = 10, height = 8, units = "in", res = 300)
    df <- simplifyGO(mat, use_raster = FALSE)
    dev.off()
  }
}
```

Full dataset line fixed effect: genes that are sex-biased on all days
```{r}
sig_list = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_figures_python/GO_enrichment/upsetplot_GO_terms_no_separation_line_random_line_ign_alldays.txt"
semantic_clistering_from_table(sig_infile=sig_list)
```

sex-separated dataset, genes that are day-biased in early (14-16) or late (16-18) comparisons in males
```{r}
sig_list = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_figures_python/GO_enrichment/upsetplot_GO_terms_sex_separated_all_line_ign_males:day14-day16.txt"
semantic_clistering_from_table(sig_infile=sig_list)
sig_list = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_figures_python/GO_enrichment/upsetplot_GO_terms_sex_separated_all_line_ign_males:day16-day18.txt"
semantic_clistering_from_table(sig_infile=sig_list)
sig_list = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/DE_figures_python/GO_enrichment/upsetplot_GO_terms_sex_separated_all_line_ign_males:day14-day18.txt"
semantic_clistering_from_table(sig_infile=sig_list)
```

Day-separated dataset, genes that are sex-biased in all days
```{r}
sig_list = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/sig_DE_genes_lists/day_separated_all_lines_sex_bias_overlap_all_days_sigIDs_sig_GO.txt"
semantic_clistering_from_table(sig_infile=sig_list)
sig_list = "/Users/miltr339/work/PhD_code/PhD_chapter4/data/sig_DE_genes_lists/day_separated_all_lines_sex_bias_overlap_day14_sigIDs_sig_GO.txt"
semantic_clistering_from_table(sig_infile=sig_list)
```